The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-09-17 conf_article/neon.tex
Cleared in the code/norm folder
Minor revision in the article

 That is what can be said from a practical point of view. From the perspective of theory, Muon's main innovation was an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update \citep{bernstein2025deriving}. Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

 In this article, we focus on the two most common matrix norms akin to the spectral, namely, the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.

 Noticing that Neon and Muon are diametrical in terms of the rank of the update matrix, we bridge the space by "regularizing" them by Normalized SGD, which is derived in lmo with the Frobenius norm. We do this in the same lmo approach by considering a norm that is dual to the convex combination of the Frobenius norm and the spectral or the nuclear norm respectively. So we derive the algorithms we name F-Neon and F-Muon respectively.

 Having faced the array of different Muon-like optimizers, according to the upper bounds from \citet{kovalev2025understanding,riabinin2025gluon}, with similar convergence behavior, we painstakingly compare them on a synthetic linear least squares problem with known Lipschitz constant. The efforts results in comparison of the algorithms by their convergence not in terms of the dual norms of their gradient, but in terms of the common spectral norm, which may strongly differ, especially in large matrices, from the initial dual norms. Thus, we compare the algorithms in a unified fashion and find that if an algorithm converges fast by the loss, it does not mean it converges fast by the norm of the gradient, and vice versa.

 Finally, we test Muon and F-Muon on two deep-learning tasks, training convolutional network on CIFAR-10 \citep{cifar2023airbench} and pre-training NanoGPT on FineWeb \citep{modded_nanogpt_2024}. The results support the supremacy of Muon, but the most striking result of the tests is that F-Muon, only half of which is Muon, surpasses Muon's accuracy on the CIFAR tasks by a margin. The case of F-Muon answers in the affirmative to the question of feasibility of constructing a mixture of optimization algorithms to increase robustness of the composite algorithm.

@@

        \begin{lemma}\label{lemma:ky_fan_update}    
            When $\norm{\cdot} = \normkfk{\mM^k}^\dagger$, \cref{eq:our_update} turns into:

            \begin{equation}\label{eq:ky_fan_update}
                \mX^{k+1} = \mX^{k} - \eta \sum_{i=1}^{k}u_i v_i^\top
            \end{equation}
        \end{lemma}
        In the section {\it The Matrix side of updates}, we will discuss how to compute the updates efficiently.

        We introduce the family of {\it Fanions}, which consists of {\it $k$-Fanions}, lmo-based algorithms under the $\normkfk{\mM^k}^\dagger$ norms. By this terminology and the lemma, Muon is a $\min\{n,m\}$-Fanion, while Neon is an 1-Fanion. Moreover, the centralized version of rank-$r$ Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback and without scaling of the update, from the perspective of lmo, is actually a $r$-Fanion.

        In addition, one can consider F-KF-k-norm = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback. The resulting family can be named {\it F-Fanions}.

@@

    As the norms $\normf{\cdot}$, $\normn{\cdot}$, $\normf{\cdot}$ are almost proportional to each other when $m, n \rightarrow \infty$ (with high probability for random matrices), the expected convergence guarantees in terms of $\normf{\cdot}$ are the same (it can be easily shown by noting that $\norm{\mX}\sim\alpha\normf{\mX}$, $\norm{\nabla f(\mX)}^\dagger \sim \frac{1}{\alpha} \normf{\nabla f(\mX)}$, and expressing $L$-constant via $L_F$-constant for the Frobenius norm).

@@

\section{Conclusion}
In this article, we have generalized several successful algorithms, like Muon and Dion, to the lmo-based algorithms in the $\normkfk{\cdot}^\dagger$ norm. Also we have proposed the technique of ``regularizing'' the updates with NSGD, a trick to increase the robustness of the algorithms and motivated by the consideration of the $\normfstar{\cdot}$ and $\normftwo{\cdot}$ norms. Generalizations of well-known norms and subsequent combination of them may further improve performance of lmo-based algorithms. If a theory is developed that explains the discrepancy between performance of different algorithms based on the matrix norms, one will probably be able to apply it to the generalizations and combinations of the norms as well, which leaves an open space to construct norms specific for given architectures and probably even their parameters.

@@

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{fstardualball.pdf}
        \caption{lmo balls for F-Muon for different $\alpha$}
        \label{fig:fstardual}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{ftwodualball.pdf}
        \caption{lmo ball for F-Neon for different $\alpha$}
        \label{fig:ftwodual}
    \end{subfigure}
    \caption{Balls in the duals to F* and F2 norms for different $\alpha$}
    \label{fig:fduals}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.45\linewidth}
        \centering
        \includegraphics[width=\linewidth]{KyFan.pdf}
        \caption{Ky Fan 2-norm}
        \label{fig:kyfan}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\linewidth}
        \centering
        \includegraphics[width=\linewidth]{KyFanDual.pdf}
        \caption{Dual to Ky Fan 2-norm}
        \label{fig:kyfandual}
    \end{subfigure}
    \caption{Ky Fan 2-norm and its dual}
    \label{fig:kyfan_combined}
\end{figure}

@@

Proof of \cref{lemma:ky_fan_update}:
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal to reach $\normkfk{\mM^k}$.

Let us note that $\Delta = \sum_{i=1}^{k} u_i v_i^\top$ delivers the value. Indeed, $\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} u_i v_i^\top> = \sum_{i,j=1}^{r, k}\<u_i \sigma_i v_i^\top, u_j v_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^k}$, which completes the proof.

>>>>

 That is what can be said from a practical point of view. From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update \citep{bernstein2025deriving}. Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

 In this article, we focus on the two most common matrix norms akin to the spectral norm, namely the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.

 Noticing that Neon and Muon are diametrical in terms of the rank of the update matrix, we bridge the space by considering the dual to the Ky Fan $k$-norm, the primal norm being the sum of the $k$ largest singular values of the matrix. The resulting update turns out to be Dion \citet{ahn2025dioncommunicationefficientoptimizerlarge} without error feedback.

 To make Neon and Dion without error feedback more robust, we "regularize" them by Normalized SGD (NSGD), which is derived in lmo with the Frobenius norm. We do this in the same lmo approach by considering a norm that is dual to the convex combination of the Frobenius norm and the Ky Fan k-norm. So we derive the algorithms we name F-Neon and F-Dion respectively. The same procedure we follow with Muon to obtain F-Muon.

 Having faced the array of different Muon-like optimizers, which, according to the upper bounds from \citet{kovalev2025understanding,riabinin2025gluon}, must converge similarly, we painstakingly compare them on a synthetic linear least squares problem with a known Lipschitz constant. The efforts result in comparison of the algorithms by their convergence not in terms of the dual norms of the gradient, but in terms of the common spectral norm, which may strongly differ, especially in large matrices, from the initial dual norms. Thus, we compare the algorithms in a unified fashion and find that sometimes if an algorithm converges fast by the loss, it does not imply its fast convergence by the norm of the gradient, and vice versa.

 Finally, observing the poor performace of Neon and Dion without error feedback, which negatively impacts F-Neon and F-Dion as well, we test Muon and F-Muon on two deep-learning tasks, training convolutional network on CIFAR-10 \citep{cifar2023airbench} and pre-training NanoGPT on FineWeb \citep{modded_nanogpt_2024}. The results support the supremacy of Muon, but the most striking outcome of the tests is that F-Muon, only half of which is Muon, is on par with Muon on training the convolutional network and only slighter worse that Muon on NanoGPT pre-training, while NSGD is far from that. The case of F-Muon answers in the affirmative to the question of possibility to perturb the norm that underpins the algorithm without sacrificing its performace. We envision that this effect will be used in the future to constuct easier to compute approximations of empirically optimal algorithms like Muon.

@@

        \begin{lemma}\label{lemma:ky_fan_update}    
            When $\norm{\cdot} = \normkfk{\mM^t}^\dagger$, \cref{eq:our_update} turns into:

            \begin{equation}\label{eq:ky_fan_update}
                \mX^{t+1} = \mX^{t} - \eta \sum_{i=1}^{k}u_i v_i^\top
            \end{equation}
        \end{lemma}
        In the section {\it The Matrix side of updates}, we will discuss how to compute the updates efficiently.

        We introduce the family of {\it Fanions}, which consists of {\it $k$-Fanions}, lmo-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. By this terminology and the lemma, Muon is a $\min\{n,m\}$-Fanion, while Neon is an 1-Fanion. Moreover, the centralized version of rank-$r$ Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback and without scaling of the update, from the perspective of lmo, is actually a $r$-Fanion.

        In addition, one can consider F-KF-k-norm: $\norm{\cdot}_{\mathrm{F-KF-k}}$ = $\alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}$, the dual to which will produce algorithms like F-Dion without the error feedback. The resulting family can be named {\it F-Fanions}.

@@

    As the norms $\normf{\cdot}$, $\normn{\cdot}$, $\norms{\cdot}$ are almost proportional to each other when $m, n \rightarrow \infty$ (with high probability for random matrices), the expected convergence guarantees in terms of $\normf{\cdot}$ are the same (it can be easily shown by noting that $\norm{\mX}\sim\alpha\normf{\mX}$, $\norm{\nabla f(\mX)}^\dagger \sim \frac{1}{\alpha} \normf{\nabla f(\mX)}$, and expressing $L$-constant via $L_F$-constant for the Frobenius norm).

@@

\section{Conclusion}
In this article, we have generalized several successful algorithms, like Muon and Dion, to the lmo-based algorithms in the $\normkfk{\cdot}^\dagger$ norm. Also we have proposed the technique of ``regularizing'' the updates with NSGD, a trick to increase the robustness of the algorithms, which is motivated by the consideration of the $\normfstar{\cdot}$, $\normftwo{\cdot}$, and general $\norm{\cdot}_{\mathrm{F-KF-k}}$ norms. Generalizations of well-known norms and subsequent combinations of them may further improve performance of lmo-based algorithms. If a theory is developed that explains the discrepancy between performance of different algorithms based on the matrix norms, one will probably be able to intentionally construct norms optimal to given architectures and probably even their parameters.

@@

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/fstardualball.pdf}
        \caption{lmo balls for F-Muon for different $\alpha$}
        \label{fig:fstardual}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/ftwodualball.pdf}
        \caption{lmo ball for F-Neon for different $\alpha$}
        \label{fig:ftwodual}
    \end{subfigure}
    \caption{Balls in the duals to F* and F2 norms for different $\alpha$}
    \label{fig:fduals}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.45\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/KyFan.pdf}
        \caption{Ky Fan 2-norm}
        \label{fig:kyfan}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/KyFanDual.pdf}
        \caption{Dual to Ky Fan 2-norm}
        \label{fig:kyfandual}
    \end{subfigure}
    \caption{Ky Fan 2-norm and its dual}
    \label{fig:kyfan_combined}
\end{figure}

@@

Proof of \cref{lemma:ky_fan_update}:
Since $\norm{\cdot}^\dagger = \normkfk{\cdot}^{\dagger\dagger} = \normkfk{\cdot}$, the goal to reach $\normkfk{\mM^t}$.

Let us note that $\Delta = \sum_{i=1}^{k} u_i v_i^\top$ delivers the value. Indeed, $\<\mM^t, \Delta> =\<\mU \mSigma \mV^\top, \sum_{i=1}^{k} u_i v_i^\top> = \sum_{i,j=1}^{r, k}\<u_i \sigma_i v_i^\top, u_j v_j^\top> = \sum_{i=1}^{k}\sigma_i = \normkfk{\mM^t}$, which completes the proof.
